---
title: "Data Analysis and Figures"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

### Functions load 
```{r}
library(tidyverse) # Data load/manipulation and graphics build
library(RColorBrewer) # Color Palettes
library(lubridate) # Day/Time Manipulation
library(patchwork) # Plot Layout
library(png) # Import `png` images
```

### Data Load
```{r}
data_oan <- read_csv("2.Datos/working_data/bc2021_data.csv")
glimpse(data_oan)
```

Define de row names that belongs to ID of data like: `date`, `site`, `river`...

```{r}
id_vars <- c("date","date_time", "estacion","river","data_model")

```

# Rename variables for easy coding
```{r}
data_oan <- data_oan %>% 
  rename(chla = "clorofila_a_mg_l",
         alk = "alcalinidad_total_mg_ca_co3_l",
         cond = "conductividad_m_s_cm",
         tot_phos = "fosforo_total_mg_p_l",
         tss = "solidos_suspendidos_totales_mg_l",
         pH = "potencial_de_hidrogeno_p_h_sin_unid",
         temp = "temperatura_o_c")
```


### 99.5 percentile limits for Chl-a

We need to assume that BC2021 calculated the `99.5 limit` using Negro and Uruguay together. Lets see how many data exceed the limit with this criteria
```{r}
data_oan %>% 
  filter(river %in% c("Negro","Uruguay")) %>% 
  filter(chla >= quantile(chla, probs = c(0.995), na.rm = T)) %>%
  dplyr::select(date, estacion, chla) %>% kable()
```
* Looks similar that they did, but with this criteria the RN12 (2018-04-17) does not get excluded
What happens if `chla` was calculated for each river?. How many values should be eliminated?

```{r}
#Function that calculates the number of data points that exceed 99.5 limit for a numeric variable 
  q99.5_exceed <- function (x){
    q <- quantile(x, probs = c(0.995), na.rm =T)
  sum(x > q, na.rm = T)
    }

```

## Number of data exceeds 99.5 for each variable in Uruguay and Negro

```{r}
# Apply the function q99.5_exceed for Uruguay and Negro
 data_oan %>% filter(river %in% c("Uruguay", "Negro"))  %>% 
   summarize(across(where(is.numeric), q99.5_exceed)) %>% 
  kable()
```
  

* Group by River and calculate the number of data exceeds 99.5 for each variable

```{r}
# Apply the function q99.5_exceed for each river separated
data_oan %>% group_by(river)  %>% 
summarize(across(where(is.numeric), q99.5_exceed)) %>% 
  kable()
```

### Calculate Q99.5 values of chla by river
```{r}
data_oan %>% group_by(river) %>% 
  dplyr::select(date, estacion, chla, river) %>% 
  summarise(chla_99.5 = quantile(chla, probs = c(0.995), na.rm = T)) %>% kable()
```

### Lets calculate de Q99.5 for any numeric variable
```{r}
# Function that returns de value of quantile 99.5  
q_calc<- function(x) {
 q <- quantile(x, probs = c(0.995), names = F,na.rm = T)
}

```

Apply de q_calc for each numeric column from Negro and Uruguay rivers

```{r}
data_oan %>% 
    filter(river %in% c("Negro","Uruguay")) %>% 
    summarise( across(where(is.numeric), q_calc)) %>% 
  kable()

```
  
### Chla vs All all values  
```{r}
# Generate de data frame to make one plot for each variable
# With a data-list each var transform into single data frame to make the plot
chla_allvalues <- data_oan %>%
  filter (river %in% c("Negro", "Uruguay")) %>% 
    pivot_longer(
    cols = !c(all_of(id_vars), chla)) %>% 
  group_by(name) %>% 
  nest()

# Function that make a plot for each variable 
figa1_func <- function (data) { 
  # Define de q99.5 limit  
q <- quantile(data$value, probs = c(0.995), names = F,na.rm = T)
  data %>% 
    mutate(extra = ifelse(value > q, "1","0") ) %>% 
  ggplot(aes(x = value , y = chla, fill = extra)) +
    geom_point(alpha = 0.8, size = 1.5, pch = 21) +
    scale_y_continuous(limits = c(0,60)) +
    scale_fill_manual(na.translate = FALSE , values = c("#984ea3", "#66a61e")) +
    labs(y = expression(paste("Chlorophyll a (", mu,"g L"^-1,")"))) +
    theme(axis.title.y = element_text(size = 8)) +
    guides (fill = "none")
         }
# Apply figa1_function for each variable

data_chla_all <- chla_allvalues %>% 
  mutate(plots = map(data, figa1_func)) 

# Extract the plots information obly 
chla_plots<- data_chla_all %>%  ungroup() %>% 
  dplyr::select(plots) 

# Extract each plot for label variables adequately

figa1_alk <- chla_plots[[1]][[1]] +
  labs(x = expression(paste("Alkalinity (mg L"^-1,")")) )

figa1_ec <- chla_plots[[1]][[2]] +
  labs(x = expression(paste("EC"[w] ," (", mu,"S cm"^-1,")")) )

figa1_tp <- chla_plots[[1]][[3]] +
  labs(x = expression(paste("Total Phosphorus (", mu,"g L"^-1,")")) )

figa1_sst <- chla_plots[[1]][[4]] +
  labs(x = expression(paste("Total Suspended Solids ( mg L"^-1,")")) ) 

figa1_ph <- chla_plots[[1]][[5]] +
  labs(x = "pH" )

figa1_ta <- chla_plots[[1]][[6]] +
  labs(x = expression(paste("T (", degree,"C)")) ) 
```

```{r figureA1,fig.align= 'center', fig.retina=2, fig.width= 9.35, fig.height= 6.94 }
wrap_plots( figa1_alk,figa1_ec,
                     figa1_tp,figa1_sst,figa1_ph,
                     figa1_ta, ncol = 2) +
  plot_annotation(tag_levels = 'a') &
  theme(plot.tag.position = c(0, 1),
        plot.tag = element_text(size = 12, hjust = 0, vjust = 0))

```
Figure 3 from BC2021 replicated with the outliers highlighted in green according our analysis (values higher than the 99.5% percentile.

### Substitute all values that exceed 99.5 and replace by NA
 * For Each variable in  `Negro` and `Uruguay` together (for similarity with the BC2021 procedure)
 * For `Cuareim` this is done alone

```{r}
# Function that replace a values with NaN if it exceed 99.5 limit

q99.5_remove <- function(x){
  q = quantile(x, probs = c(0.995), na.rm = T)
  x_c = ifelse(x <= q , x , NaN)
}

# For Negro and Uruguay together

data_cut_NU <- data_oan %>% 
  filter (river %in% c("Negro","Uruguay")) %>% 
  mutate(across(where(is.numeric), q99.5_remove))

# For Cuareim only
data_cut_C <- data_oan %>% 
  filter (river == "Cuareim") %>% 
  mutate(across(where(is.numeric), q99.5_remove))

bc_data_limit <- bind_rows(data_cut_NU,data_cut_C) 

```

## chla vs environment  
### Figure 3 from BC 2021 recreation

Store the maximum values for X axes according to BC2021(Figure 3)
```{r}
plot_x_limits <- tribble(
  ~name, ~lmin, ~lmax,
  "alk", 0, 150,
  "cond",0, 300,
  "tot_phos", 0, 1000,
  "pH", 5,10,
  "tss", 0, 600,
  "temp", 0 , 40
)

```


Generate de data frame to make one plot for each variable. With a data-list each var transform into single data frame to make the plot

```{r}
data_fig3 <- bc_data_limit %>% 
  filter (river != "Cuareim") %>% 
  dplyr::select(!all_of(id_vars)) %>% 
  pivot_longer (cols =! chla) %>% 
  group_by (name) %>% 
  nest()

# Add x limits
data_fig3 <- data_fig3 %>% 
  left_join(plot_x_limits, id = "name")
```


```{r}
# Function that generates a plot for any numerical variable vs chla
# Set x axes range according to the same as BC2021
fig3_function_plot <- function (data, lmin,lmax,xlab) {
  ggplot(data, aes(x = value , y = chla)) +
    geom_point(alpha = 0.8, size = rel(0.5)) +
    scale_x_continuous(limits = c( xmin = lmin , xmax = lmax)) +
    scale_y_continuous(limits = c(0,45),
                       breaks = seq(from = 0,to=48,by = 5)) +
    labs( x = xlab, y = expression(paste("Chlorophyll a (", mu,"g L"^-1,")"))) +
    theme_classic() +
    theme(axis.title = element_text(size = rel(0.7)),
           plot.margin = unit(c(0, 0, 0, 0), "cm"))
    
}

# Apply the plot function for each variable

data_fig3 <- data_fig3 %>% 
  mutate (plot = pmap(list(data,lmin,lmax,name), fig3_function_plot))

# Extract only plots information
plots3<- data_fig3 %>%  ungroup() %>% 
  dplyr::select(plot) 

```

Extract each plot for label variables adequately
```{r}
fig3_alk <- plots3[[1]][[1]] +
  labs(x = expression(paste("Alkalinity (mg L"^-1,")")) ) +
   coord_fixed(ratio = 3)

fig3_ec <- plots3[[1]][[2]] +
  labs(x = expression(paste("EC"[w] ," (", mu,"S cm"^-1,")")) ) + 
   coord_fixed(ratio = 6)

fig3_tp <- plots3[[1]][[3]] +
  labs(x = expression(paste("Total Phosphorus (", mu,"g L"^-1,")")) ) +
 coord_fixed(ratio = 20)

fig3_sst <- plots3[[1]][[4]] +
  labs(x = expression(paste("Total Suspended Solids ( mg L"^-1,")")) ) +
   coord_fixed(ratio = 12)

fig3_ph <- plots3[[1]][[5]] +
  labs(x = "pH" ) + 
   coord_fixed(ratio = 0.1)

fig3_ta <- plots3[[1]][[6]] +
  labs(x = expression(paste("T (", degree,"C)")) ) +
   coord_fixed(ratio = 0.8)

```

### Load all panel figures copied from BC2021 that are saved each one ina differente `.png` file
```{r}
# Special case functions for import png figure and combine with ours
library(png) # To upload png images obtained from BC2021
library(grid) # Convert PNG to Grob
library(ggplotify) # Convert Grob to GGPLOT
library(cowplot) # Paste plots together

# Load `png` and transform
f3a<-readPNG("2.Datos/Fig3BC/3a.png")
fig3a <- as.ggplot(grid::rasterGrob(f3a, interpolate=TRUE)) +
    theme(plot.margin = unit(c(0, -1, 0, -1), "cm"))

f3b<-readPNG("2.Datos/Fig3BC/3b.png")
fig3b <-as.ggplot(grid::rasterGrob(f3b, interpolate=TRUE)) +
    theme(plot.margin = unit(c(0, -1, 0, -1), "cm"))

f3c<-readPNG("2.Datos/Fig3BC/3c.png")
fig3c <- as.ggplot(grid::rasterGrob(f3c, interpolate=TRUE)) + 
    theme(plot.margin = unit(c(0, -1, 0, -1), "cm"))

f3d<-readPNG("2.Datos/Fig3BC/3d.png")
fig3d <- as.ggplot(grid::rasterGrob(f3d, interpolate=TRUE)) +
    theme(plot.margin = unit(c(0, -1, 0, -1), "cm"))

f3e<-readPNG("2.Datos/Fig3BC/3e.png")
fig3e <- as.ggplot(grid::rasterGrob(f3e, interpolate=TRUE)) +
  theme(plot.margin = unit(c(0, -1, 0, -1), "cm"))

f3f<-readPNG("2.Datos/Fig3BC/3f.png")
fig3f <- as.ggplot(grid::rasterGrob(f3f, interpolate=TRUE)) +
    theme(plot.margin = unit(c(0, -1, 0, -1), "cm"))
```

### Plot figure 3 

```{r Figure 3, fig.align= 'center', fig.retina=2, fig.width= 9.35, fig.height= 6.94}
plot_grid(fig3a, fig3_alk ,
                           fig3b, fig3_ec ,
                           fig3c, fig3_tp ,
                           fig3d, fig3_ph ,
                           fig3e, fig3_sst ,
                           fig3f, fig3_ta ,
                           labels = c("", "a'",
                              "", "b'",
                              "", "c'",
                              "", "d'",
                              "", "e'",
                              "", "f'"),
                           label_x=0.8,
                           label_fontface = "plain",
                           label_fontfamily = "times", ncol = 2)
```
Biplots of Chl-a versus all environmental variables: a: alkalinity, b: EC, c: total phosphorous, d: pH, e:total suspended solids, f: temperature. In all comparative cases, the graphs redrawn from BC2021 are show to the left (letters) and our graphs to the right (prime-letters). To facilitate the visual comparison between Figure 3 of BC2021 and this paper, the axes (x and y) of the biplots are drawn with the same scaling. Note that this was not applied for the pH (in BC2021 the axis goes up to 8) and in our case that dropped 58 values on the outside(d´). The Chl-a value of 55 µg L-1 that BC2021 considered to as an outlier (RN 2018-04-17) is not shown in any of the biplots, because the Chl-a axis ends at 45 µg L-1.


### Differences within rivers used for train model (Negro and Uruguay)

Generate de data frame to plot the variables for each river. With a data-list each var transform into single data frame to make the plot

```{r}
diff_rivers_data <- bc_data_limit %>% 
   mutate(logChla = log10(chla)) %>% 
  pivot_longer (!all_of(id_vars)) %>% 
  mutate (river_label = fct_recode(river,
                                   UR = "Uruguay",
                                   NR = "Negro",
                                   CR = "Cuareim")) %>% 
  group_by(name) %>% 
  nest()

# Function that for each variable plot the violins seprating each river and facetting if data were used for model train or test

fig4_function <- function(data) {
  p1 <- ggplot(data, aes(y = value, x = river_label)) + 
    geom_violin(aes(fill = river)) +
    scale_fill_manual(na.translate = FALSE , 
                      values = c("#d95f02", "#1b9e77","#984ea3"))+
    theme(legend.position = "none",
          axis.text.x = element_text(size = rel(1.2)),
          axis.text.y = element_text(size = rel (0.7))) +
    labs(x = NULL) +
    facet_grid( ~ fct_relevel(data_model,"Train","Test"),
                scales = "free_x",
                space = "free_x")
  
}

# Apply the plot function for each variable 
fig4_rivers <- diff_rivers_data %>% 
  mutate(plots = map(data, fig4_function))

# Extract only plots information
plots4<-fig4_rivers %>%  ungroup() %>% 
  dplyr::select(plots) 

```

Extract each plot for label variables adequately

```{r}
fig4_chla <- plots4[[1]][[1]] +
  labs(y =  expression(paste("Chlorophyll a (", mu,"g L"^-1,")")) )


fig4_alk <- plots4[[1]][[2]] +
  labs(y = expression(paste("Alkalinity (mg L"^-1,")")) )

fig4_ec <- plots4[[1]][[3]] +
  labs(y = expression(paste("EC"[w] ," (", mu,"S cm"^-1,")")) )

fig4_tp <- plots4[[1]][[4]] +
  labs(y = expression(paste("Total phosphorus (", mu,"g L"^-1,")")) )

fig4_sst <- plots4[[1]][[5]] +
  labs(y = expression(paste("Total Suspended Solid ( mg L"^-1,")"))) +
  theme(axis.title.y =  element_text(size = 10))

fig4_ph <- plots4[[1]][[6]] +
  labs(y = "pH" )

fig4_ta <- plots4[[1]][[7]] +
  labs(y = expression(paste("T (", degree,"C)")) )

fig4_logchla <- plots4[[1]][[8]] +
  labs(y =  expression(paste("log"[10], "(Chlorophyll a)")) )

```


Plot figure 4
```{r Figure4, fig.align= 'center', fig.retina=2,  fig.width= 9.35, fig.height= 7.94 }
wrap_plots(fig4_chla, fig4_logchla,
                      fig4_alk,fig4_ec,
                      fig4_tp,fig4_sst,fig4_ph,
                      fig4_ta, ncol = 2)

```
Violin plots for environmental variables for Negro River (NR, green), Uruguay River (UR, violet) and Cuareim River (CR, orange) respectively. NR and UR data were used to train the model and data from CR to test model performance. UR and NR exhibit significant differences (p-values < 0.05) in mean and variance for all environmental variables, except for temperature variance. The full description of statistical test is shown in Table S4 of Supplementary Material. Note that the Cuareim River was subject to the same outlier removal procedure as explained in section 1c of the methodology

### Generalized Least Squares (GLS) for river comparision 

```{r}
#This GLS function allows for compare differences between mean and variances
library(nlme)
gls.var.test<-function(data){
  x = data$x
  group = data$group
  if(is.null(group)){ data.gls<-x; colnames(data.gls)<-c("x","group")} else data.gls<-data.frame(x,group)
  
  if(any(!is.finite(data.gls[,1]))){ data.gls<-data.gls[which(is.finite(data.gls[,1])),]; warning("there were NaNs in the original x data")}# SACA LOS NAN
  if(any(!is.finite(data.gls[,2]))){ data.gls<-data.gls[which(is.finite(data.gls[,2])),]; warning("there were NaNs in the original group data")}# SACA LOS NAN
  
  modHeteroVar = gls(x~group, data=data.gls, weights = varIdent(form = ~1|group), method="ML") # Heterogeneous variance
  modHomoVar   = gls(x~group, data=data.gls, method="ML") # Homogeneous Variance
  modEqualMean = gls(x~1,     data=data.gls, weights = varIdent(form = ~1|group), method="ML")# Same mean all groups, different variance
  gls_data <- data.frame(VarTest=anova(modHeteroVar,modHomoVar)$`p-value`[2],
                         MeanTest=anova(modEqualMean,modHeteroVar)$`p-value`[2]) # Evaluar el loglikelyhood ratio test. p>0.01
}

```

Generate the data frame necessary for compare variables between rivers, run the gls and extact information to a table


```{r}
 diff_rivers_data %>% 
  unnest(cols = data) %>% 
  filter (river != "Cuareim") %>% 
  rename(x = value ,
        group = river ) %>% 
mutate (group = factor(group) )%>% 
  nest() %>% 
  mutate(gls = map(data, gls.var.test)) %>% 
  unnest(gls) %>% 
  mutate( VarTest = ifelse( VarTest <= 0.05, "p<0.05","n.s"),
          MeanTest = ifelse( MeanTest <= 0.05, "p<0.05","n.s")) %>% 
  dplyr::select(name,VarTest, MeanTest) %>% 
  kable()

```

#### Session info

```{r}
sessionInfo()
```

