---
title: "Sites Spatail Data"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```
#### Packages

```{r}
library(readxl) # to read xls files
library(tidyverse) # Data load/manipulation and graphics build
library(janitor)# Data headers make clean
library(lubridate) # Day/Time Manipulation
library(snakecase) # Factor levels and names manipulation
library(sf) # simple features for spatial data
library(mapsf) # helps to design various cartographic representations
library(geouy) # easily access official spatial data sets of Uruguay
library(ggspatial) # Spatial data framework for ggplot2
library(viridis) # color palette

#devtools::install_github("yutannihilation/ggsflabel") 
library(ggsflabel)# Experimental packages for better labelling of sf maps
###
```

### Site Coordinates table

```{r}
sites <- read_excel("2.Datos/gis_data/sites_references.xls",
                     .name_repair = make_clean_names) %>% 
  mutate(
    programa = to_snake_case(programa),
    nombre_punto = to_snake_case (estacion),
    .keep= "unused")
```


Delete station from the bottom idetified by and `F` and the code end label
```{r}
 sites <- sites %>% filter(!str_detect(codigo_estacion,"F"))

```
 
Also load  OAN Database, to match stations and calculte freqeuncy and dates
```{r}
oan_data <- read_csv("2.Datos/working_data/OAN_complet_data.csv")

```

Filter by data points from oan data, generate the varyble type and arrange by `latitude`

```{r}
 sites_reply <- sites %>% 
  filter(codigo_estacion %in% oan_data$estacion) %>% 
     mutate(type = ifelse(str_detect(codigo_estacion, c("RN","RU")), "train", 
                       ifelse( str_detect(codigo_estacion, "RC"),"test","not_used"))) %>%
  arrange(latitud, .desc = TRUE) %>% 
   mutate(site_number = row_number())

```


### Sampling stations data

 Calculates n data per site and starting end dates

```{r}
sites_oan_data <- oan_data %>% 
   group_by(estacion) %>% 
   summarize ( N_Station = length(estacion),
               Start_Date = min(date),
               End_Date = max(date)) %>% 
   rename (Sampling_Point = "estacion")

```

#### Table A1: Sampling station
```{r}
sites_reply %>% 
   rename (Sampling_Point = "codigo_estacion",
           Latitude = "latitud",
           Longitude = "longitud",
           `Site Code` = "site_number") %>% 
    mutate (Basin = ifelse(str_detect(Sampling_Point,c("RC","RU","SSS")), "Uruguay","Negro")) %>% 
     full_join(sites_oan_data, by = "Sampling_Point") %>% 
      dplyr::select(Basin,Sampling_Point,Latitude,Longitude,type,N_Station,Start_Date,End_Date, `Site Code`) %>% 
  kable()

```
 
 
# Map With Sampling Stations 

### Convert it to spacial object with crs 4326 projection
```{r}

sites_longlat <- sites_reply %>%
  st_as_sf (coords = c("longitud", "latitud")) %>%
  st_sf (crs = 4326)
```

### Load Uruguay border cartography from package `geouy`


```{r}
borders <- load_geouy("Uruguay")

rivers <-
  load_geouy("Cursos de agua navegables y flotables") %>% # Rivers cartography
  st_cast("MULTILINESTRING") # Convert multicurve to multistring https://github.com/r-spatial/sf/issues/1194

```

### Selected rivers and streams to plot
```{r}
# A vector with river names separated by conditional symbol "|"
this_rivers <- paste(
  "uruguay",
  "cuareim",
  "río_negro",
  "tacuarembó",
  "tacuarembó_chico",
  "de_la_plata",
  "san_salvador",
  sep = "|"
)

```

Filter by selected rivers
```{r}
study_rivers <- rivers %>%
  mutate(nombre = to_snake_case(nombre)) %>%
  filter(str_detect(nombre, this_rivers))

```
### Reservoirs and lagoons cartography dowloaded from DIva-Gis Free Spatial Data at country level `http://diva-gis.org/gdata`

```{r}
water_bodies <-
  st_read ("2.Datos/gis_data/ury_inlandwater") %>%  # Uruguay/InlandWater
  mutate (name = to_snake_case(NAME),
          .keep = "unused") %>%  # te drop de old NAME column and retain the new one
  filter(name %in% c("rio_uruguay_uruguay", "embalse_del_rio_negro"))

```

### Generate Base Map

```{r}
base_map <- ggplot() +
  geom_sf (data = borders, fill = "NA", color = "grey40") +
  geom_sf (data = study_rivers, color = "#1f78b4") +
  geom_sf (data = water_bodies, color = "#1f78b4") +
  theme_bw() +
  theme(panel.grid = element_blank())

```

```{r}
# Edit Rivers names, add to the plot maintaining its x,y coordinates

river_labels <- study_rivers %>%
  mutate(
    name = fct_recode(
      nombre,
      `Tbo Chico river` = "arroyo_tacuarembó_chico",
      `Negro river` = "río_negro",
      `Uruguay river` = "río_uruguay",
      `Cuareim river` = "río_cuareim",
      `San Salvador river` = "río_san_salvador",
      `Tbo river` = "río_tacuarembó",
      `Río de la Plata` = "río_de_la_plata"
    )
  )

```


```{r}
# Map with site points filled if are used or not into BC2011
mapa_listo <-  base_map +
  geom_sf(
    data = sites_longlat,
    aes(fill = type, shape = type),
    size = 1.5,
    color = "black"
  ) +
  scale_fill_viridis(option = "magma", discrete = T) +
  scale_shape_manual(values = c(21, 22, 23), guide = "legend") +
  labs (fill = "Legend", shape = "Legend",) +
  annotation_scale(location = "br",
                   width_hint = 0.2,
                   text_cex = 0.5) +
  annotation_north_arrow(
    location = "bl",
    which_north = "true",
    pad_x = unit(3, "in"),
    pad_y = unit(3.5, "in"),
    style = north_arrow_fancy_orienteering
  ) +
  labs (x = "longitude", y = "latitude") +
  geom_sf_text(
    data = river_labels,
    aes(label = name),
    size = 3,
    fontface = "bold",
    family = "serif",
    hjust = c(
      `Tbo Chico river` = .7,
      `Negro river` =  .2,
      `Uruguay river` = 1,
      `Cuareim river` =  .3,
      `San Salvador river` = -.5,
      `Tbo river` = .3,
      `Río de la Plata` = 0.8
    ) ,
    vjust = c(
      `Tbo Chico river` = 1.8,
      `Negro river` =  3,
      `Uruguay river` = 1.7,
      `Cuareim river` =  -1.2,
      `San Salvador river` = 0,
      `Tacuarembó river` = -1.8,
      `Río de la Plata` = 2
    ),
    angle = c(
      `Tbo Chico river` = 310,
      `Negro river` =  0,
      `Uruguay river` = 260,
      `Cuareim river` =  320,
      `San Salvador river` = 340,
      `Tacuarembó river` = 300,
      `Río de la Plata` = 0
    )
  )


```

```{r}
# Generate a character with equivalencies between number site and its code
caption_sites <- paste(sites_longlat$site_number,
                       sites_longlat$codigo_estacion,
                       sep = " = ") %>%
  str_c(., collapse = ", ")

```
# Published Map

```{r,fig.align= 'center', fig.retina=2, fig.width= 9.35}
 mapa_listo +
  geom_sf_text_repel(
    data = sites_longlat,
    aes(label = site_number),
    size = 2,
    max.overlaps = 20,
    box.padding = unit(0.1, "lines")
  )

```

```{r}
sessionInfo()
```
 
 